---
title: "Effect of rest on soccer and tennis match outcomes"
author: "Chloé Lepert"
date: "8/9/2018"
output: pdf_document
header-includes:
- \usepackage{float}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, warning = FALSE, message = FALSE)
library(ggplot2); theme_set(theme_bw())
```

# 1. Introduction

Many factors influence athletic performance in competitions. Most models aiming to predict results focus on the strength of the team and location of the competition, but other factors that players and teams can control could have an impact. In this paper we will look at how rest and tiredness impact the performance of soccer teams and tennis players. 

## 1.1 Question 

We aim to quantify the effect of rest on Football match outcomes. We ask two questions looking at short and long term rest levels of a team:

\begin{enumerate}
    \item How does the number of days since a previous match impact the performance of a team?
    \item How the the game load of a team in the past month, 2 month, and 3 month impact the performance of a team?
    
\end{enumerate}

We will look at different models in which performance will either be the probability of winning, drawing, and losing or the expected number of goals scored.

## 1.2 English Football

Before explaining our motivation we will first given an overview of the English and European soccer system so that the reader may better understand our motivation. 

English soccer is comprised of multiple competitions. 

\begin{itemize}
    \item The top 20 teams in England play in the \textbf{Premier League}. They will play each other time once at home and once away resultung in 380 games played each season. Each match earns teams points; a win will get a team 3 points, a draw one point, and loss 0 points. The team with the most points win. The bottom two teams will be relegated to the EFL Championship. The top two teams from EFL Championship will advance to the Premier league. The third to last team in the Premier league and third team in the EFL Championship will play off to play in the Premier league. \cite{PremierLeague}
    \item The \textbf{FA cup} is a knockout tournament open to teams in 10 levels of English football, this a knockout tournament. Teams enter the tournament at different times depending on the level they play in. Teams that play in the Premier league will enter the 3rd round in January with the hope of making it to the final played in May. \cite{FACup}
    \item The \textbf{EFL cup} is a tournament similar to the FA cup but open only to levels 1 through 4. Premier league teams will enter in August. Matches continue until the final in February. Teams playing in European competitions do not play in this cup \cite{EFLCup}
\end{itemize}

There are two European competitions which are played with teams from different European leagues:

\begin{itemize}
    \item The top teams play in the \textbf{Champions League}. It starts with a group stage of 32 teams in which mini round robins are run in each group to determine which 16 teams go on to the tournament. Teams not advancing to the tournament are transferred to the Europa league. In the tournament stage teams meet twice: once home and once away. Since only one team can move forward to the next round, the return leg can go overtime to decide who win. The final is played in one match in a location determined ahead of time. \cite{Champions} In England the top three premier league teams automatically qualify for the Champions league. The fourth place team qualifies for a play-off round. \cite{PL}
    \item The \textbf{Europa league} is similar to the Champions league in format but played with the top teams not qualifying for the champions league. Depending on how a team qualifies it will enter the competition at different rounds. \cite{EuropaLeague} The 5th place Premier league team qualifies for the group stage, with the winners of the FA cup and EFL cup qualifying for earlier rounds. \cite{PL}
\end{itemize}

Depending on the year, the schedule will be inter-spaced with 2-week \textbf{international breaks}, in which players are called to represent their national team in international friendlies or qualifying matches for continent or world cups. 

The wide variety of competition that a team could play in means that teams can have different schedule load as shown in the following figure.

```{r}
source("~/Documents/masters_paper/Figures/Make/gameLoad.R")
p
```

## 1.2 Relevance

Football game scheduling varies from league to league. In particular, some leagues take a winter break while others do not. Some speculate that this winter break influences performance in inter-league competitions such as the Champions and Europa league by allowing teams that have the break to be more refreshed. \cite{Winter}

Further Champions and Europa league games are played Tuesday through Thursday, while club games are played Friday through Mondays, leading to a variety of rest days going into Champions and Europa league games. (NEED CITATION) Some leagues are more willing to accommodate teams playing in European competitions than others by letting them schedule their games earlier in the weekend.

## 1.3 Approach

While we would ideally want to look at how rest affects performance in european competitions doing so is difficult because the teams that play each other in European competitions do not do so enough for us to calculate and control for team ability on an even scale. We will instead model the outcome of Premier League matches, which are set up such that we can easily control for team ability. We will evaluate how matches in competitions other than the premier league influence premier league matches.

This approach assumes that European competition games are similar Premier League games. This is mostly true except for the fact that European competition games can result in overtime and are higher stakes; a loss results in elimination. We assume that if we see rest impact outcomes in premier league games, rest is likely to have an impact in European games. 


## 1.4 Literature review

### 1.4.1 Rest time

Most of research on impacts of game scheduling focuses on injury and measures of certain activities in games (distance ran, number of sprints, etc.), but not on overall game outcomes. 

Over the past decades, Carlos Lago Penas conducted a series of study looking at how physical behaviors of players such as distance run at certain speeds evolve over a series of consecutive games with small rest periods and found minor to no differences in physical behaviors across games. \cite{Penas1, Penas2}

In 2010, Dupont et al. found a higher injury rates for football players who played two matches a week compared to players playing 1 match a week.

### 1.4.2 Modeling match outcomes

In 1982, M. J. Maher introduced 2 independent Poisson distributions as a way to model soccer scores. He proposed using a team's attacking strength and its oponent's defensive weakness as predictors of number of goals scores. Maher also found that using a Bivariate Poisson distribution with a correlation of 0.2 improved his model's fit. \cite{Maher}

In 2003, Karlis and Ntzoufras proposed a diagonal inflated bivariate Poisson model in which the probabilities of draws are increased. \cite{karlis} They also created an R package to fit biavariate Poisson GLMs which we will utilize. \cite{bivpois}


## 1.5 Data

### 1.5.1 Data set

The R package "engsoccerdata" provides us with scores from matches in England as well as European competition matches. \cite{Curley} We limit ourselves to matches occurring in seasons 1995 through 2015. The design of English Football stays consistent throughout this time range and the data set is complete for these years.

### 1.5.2 Response variable

We use two different response variables: 1. The number of home and away goals and 2. Weather the a team won, tied, or lost. The distribution of home and away goals is shown bellow.

```{r include=FALSE}
source("~/Documents/masters_paper/Figures/Make/goalDist.R")
```

```{r}
plot(p)
```


Most models for the number of goals by team in a game assume he number of home and away goals follow a poisson distribution. We see that this approximately true. There tends more games than expected with zero goals and the variance is slightly larger than the mean number of goals.

A game is a home win if the number of home goals exceeds the number of visitor goals, tied if both teams score the same number of goals, and a visitor win if the number of visitor goals exceeds the number of home goals. The distribution of outcomes is shown below. 

\begin{table}[H] \centering 
  \caption{Distribution of game outcomes} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}} cccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Outcome & Number of games & Share of games \\ 
\hline \\[-1.8ex] 
Home win & $1,966$ & $0.458$ \\ 
Tie & $1,143$ & $0.266$ \\ 
Visitor win & $1,187$ & $0.276$ \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 

### 1.5.3 Predictors

Using this data set we calculate the number of days since the previous game. Before performing out analysis on the effect of rest time we remove games in which either of the team's previous game was over 8 days ago. If team's last non-international game was more than 8 days prior, it is possible that an international break previous happened which we have no way of controlling for. 

The distribution of rest time for the home and away team is shown below. 

\begin{table}[H] \centering 
  \caption{Distribution of rest time (in days) for home and away teams} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}} ccccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 &  &  &  &  & Visitor &  &  &  \\ 
 &  & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
\hline \\[-1.8ex] 
 & 2 & $128$ & $9$ & $1$ & $0$ & $3$ & $0$ & $1$ \\ 
 & 3 & $12$ & $343$ & $155$ & $22$ & $38$ & $121$ & $7$ \\ 
 & 4 & $0$ & $165$ & $238$ & $43$ & $38$ & $91$ & $61$ \\ 
Home & 5 & $1$ & $26$ & $50$ & $63$ & $31$ & $90$ & $31$ \\ 
 & 6 & $3$ & $22$ & $34$ & $42$ & $110$ & $278$ & $41$ \\ 
 & 7 & $0$ & $118$ & $74$ & $93$ & $302$ & $798$ & $110$ \\ 
 & 8 & $1$ & $3$ & $47$ & $42$ & $32$ & $106$ & $272$ \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 

We also create a binary variable for weather or not a team is "more" (7-8 days) or "less" (1-6 days) rested and find the following distribution of rest. 

\begin{table}[H] \centering 
  \caption{Distribution of rest time for home and away teams} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}} cccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 &  & Away &  \\ 
 &  & 1-6 days & 7-8 days \\ 
\hline \\[-1.8ex] 
Home & 1-6 days & $1,577$ & $721$ \\ 
 & 7-8 days & $712$ & $1,286$ \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 

### 1.5.4 Control variables

Using Premier games we calculate a teams attacking and defensive weakness. \cite{Cronin} The attacking strength is the ratio of the average number of goals score by a team in a season to the average number of goals scored in the league that season. The defensive weakness is the ratio of the average number of goal conceded by a team in a season to the average number of goals conceded in the league that season. For each team the attacking strength and defensive weakness are computed for home and away games. The better a team the higher its attacking strength and the lower its defensive weakness strength will be. We will use the previous years attacking and defensive weakness to control for a team's ability. We exclude recently promoted teams from our analysis as we cannot use their previous season to calculate such strengths. The distribution of attacking and defensive strengths is shown below.

```{r, cache=FALSE}
source("~/Documents/masters_paper/Figures/Make/strengthDist.R")
p
```

We also control for the number of games a team plays each season. Better teams will last longer in in playoff competitions (FA Cup, Europa/Champions League, etc.) that happen concurrently with premier league games. These teams will play more games per season and thus on average have lower rest times. In particular, the design of the premier league in which each pair of teams plays each other once at each team's home stadium allows to easily calculate attacking strength and defensive weakness.

### 1.5.5 Matches looked at

As described in the previous subsections there are two points at which we remove premier league games from our data set.
\begin{enumerate}
  \item{Matches with recently promoted teams. Since we use the previous premier league season to calculate the team's attacking strength and defensive weakness we cannot calculate these quantities for recently promote teams. Removing these matches removes 2540 out of 7980 matches (32\%).}
  \item{Matches with over 8 days of rest. It is likely that an international match happened during the break which we do not have in our data. Leaving these matches in would lead to low rest matches presenting themselves as high rest matches. Removing these matches removes 1144 out of 5440 matches (21\%).}
\end{enumerate}

# 2 Methods

## 2.1 Poisson model

The Poisson model is most commonly used to model soccer scores. It assumes that the number of goals scored by a team in a soccer game follows a Poisson distribution of some parameter. We will use a GLM with a poisson link to predict the parameter. 

### 2.1.1 Model

\begin{itemize}
    \item Let $j={h, a}$ be an indicator for whether we are modeling home or away games.
    \item Let $G_{j,i}$ be the number of home or away goals in game $i$.
    \item Let $x_i$ be the predictors for game i.
    \item Assume that $G_j \thicksim Poisson(\lambda_j)$. $P_P(G_{j,i}=g_{j,i})=\frac{\lambda_j^{g_{j,i}} e^{\lambda_{j,i}}}{g_{j,i} !}$
    \item The parameter $\lambda_j$ is a linear combination of the predictors $X_j$: $\lambda_j = X_j \beta_j$  
\end{itemize}

### 2.1.2 Verifying the Poisson assumption under the null

One of the defining features of the Poisson model is the fact that the mean of a Poisson distribution equals it's variance. As the figure below shows for most teams this equality holds. 

```{r}
source("~/Documents/masters_paper/Figures/Make/goalMeanVsVariance.R")
p
```


## 2.2 Linear model for goal difference

The square root of a poisson distribution can be approximated by a nornal distribution (cite). We can therefore approximate the difference in square root goals as the difference of two normal distributions, which is itself a normal distribution. This allows us to fit a normal model to the transformed goal difference. The figure below shows the distribution of normal goal difference and its approximation as a normal distribution.

```{r, cache=FALSE}
#http://tina.wiau.man.ac.uk/docs/memos/2001-010.pdf
source("~/Documents/masters_paper/models/lmGoalDiff.R")
p
```

Our model aims to find a linear combination of our predictors that best approximates the square root goal difference

$\sqrt{G_{h,i}} - \sqrt{G_{v,i}} = X_j \beta_j$

In this model we could use difference in the number of rest days as a predictor instead of number of rest days for the home team and number of rest days for the away team. Using the difference in rest as a predictor is equivalent to using the the rest days for the home team and away teams and forcing their coefficients in the model to be the opposite. We chose to not add this constraints and allow home and away rest to vary freely.

# Results

## Simple Poisson GLM

Table \ref{table:glm_simple} summarizes the GLM models for the number of home and away goals. Rest comes in the model either as number of days since the previous game (1) and (3) or as whether or not the team had more than 6 days of rest (2) and (4). In neither of the 4 models do we see rest as having an effect on the number of goals scored. 

The strongest effects come from the attacking strength of the home team and defensive weakness of the visitor team. This not to surprising as most models looking to predict soccer outcomes use the product of these two quantities as the expected number of goals. A team with a high attacking strength should score more since it has in the past, and a team with a high defensive weakness should take in more goals as it has in the past.

We also see an effect from the game load of a team. Game load also measure the quality of a team. Better teams will qualify for more games and thus have a higher game load. We expect to see an increase in a team's game load increase its predicted number of goals and an increase in its opponent's game load decrease a team's predicted number of goal; which is what we see.

\begin{table}[H] \centering 
  \caption{Generalized Linear Models with Poisson link} 
  \label{table:glm_simple} 
\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{2}{c}{hgoal} & \multicolumn{2}{c}{vgoal} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Team\_rest & $-$0.004 &  & $-$0.007 &  \\ 
  & (0.009) &  & (0.011) &  \\ 
  Opp\_rest & $-$0.001 &  & 0.005 &  \\ 
  & (0.009) &  & (0.011) &  \\ 
  Team\_rest\_bin &  & $-$0.007 &  & $-$0.029 \\ 
  &  & (0.028) &  & (0.032) \\ 
  Opp\_rest\_bin &  & $-$0.019 &  & 0.013 \\ 
  &  & (0.027) &  & (0.032) \\ 
  Team\_att\_str & 0.371$^{***}$ & 0.372$^{***}$ & 0.427$^{***}$ & 0.427$^{***}$ \\ 
  & (0.050) & (0.050) & (0.053) & (0.053) \\ 
  Opp\_def\_weak & 0.274$^{***}$ & 0.277$^{***}$ & 0.238$^{***}$ & 0.239$^{***}$ \\ 
  & (0.061) & (0.061) & (0.061) & (0.061) \\ 
  Team\_load & 0.015$^{***}$ & 0.016$^{***}$ & 0.011$^{***}$ & 0.011$^{***}$ \\ 
  & (0.003) & (0.003) & (0.003) & (0.003) \\ 
  Opp\_load & $-$0.020$^{***}$ & $-$0.021$^{***}$ & $-$0.024$^{***}$ & $-$0.024$^{***}$ \\ 
  & (0.003) & (0.003) & (0.004) & (0.003) \\ 
  Constant & $-$0.036 & $-$0.044 & $-$0.025 & $-$0.021 \\ 
  & (0.205) & (0.198) & (0.231) & (0.223) \\ 
 \hline \\[-1.8ex] 
Observations & 4,296 & 4,296 & 4,296 & 4,296 \\ 
Log Likelihood & $-$6,597.132 & $-$6,596.946 & $-$5,804.871 & $-$5,804.693 \\ 
Akaike Inf. Crit. & 13,208.260 & 13,207.890 & 11,623.740 & 11,623.390 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 


## Linear model for goal difference

Table \ref{table:lm_simple} summarizes the model for the square root goal difference. We find that the rest of the home and away team have no significant impact on the goal difference. Predictors highlighting the strength of the teams involved are most important.

\begin{table}[H] \centering 
  \caption{Linear model for the difference in goals scored} 
  \label{table:lm_simple} 
\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} 
\\[-1.8ex] & goal\_diff \\ 
\hline \\[-1.8ex] 
 h\_rest & $-$0.005 \\ 
  & (0.010) \\ 
  v\_rest & $-$0.003 \\ 
  & (0.010) \\ 
  h\_att\_str & 0.399$^{***}$ \\ 
  & (0.057) \\ 
  h\_def\_weak & $-$0.277$^{***}$ \\ 
  & (0.057) \\ 
  v\_att\_str & $-$0.380$^{***}$ \\ 
  & (0.050) \\ 
  v\_def\_weak & 0.364$^{***}$ \\ 
  & (0.066) \\ 
  h\_load & 0.019$^{***}$ \\ 
  & (0.003) \\ 
  v\_load & $-$0.015$^{***}$ \\ 
  & (0.003) \\ 
  Constant & $-$0.037 \\ 
  & (0.246) \\ 
 \hline \\[-1.8ex] 
Observations & 4,296 \\ 
R$^{2}$ & 0.138 \\ 
Adjusted R$^{2}$ & 0.136 \\ 
Residual Std. Error & 0.878 (df = 4287) \\ 
F Statistic & 85.434$^{***}$ (df = 8; 4287) \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

# Discusion

## Importance of game load

# Conclusion

# Appendix

# References

# Notes

* in 1.1 Do we still care about long term?
* 2.1.2 Remake the figure

* Fix all the numbering